Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-23 paper.tex
update from Bara

The reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates} (see Fig.~\ref{fig::motiv}).
The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand.
This can help them to reduce the number of necessary in-vitro experiments.
The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), 
i.e., they approximate the ligand by a bounding sphere.
However, ligands are typically of a non-spherical shape which is completely ignored by these methods.

@@

Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
This necessarily requires to consider also conformational changes of the ligand, i.e., the mutual position of the ligand atoms can change.
Such a ligand is denoted as flexible ligand.
%Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
%\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
%Such a ligand, for which the mutual position of its atoms can change, is denoted as a flexible ligand.

@@

Finding trajectories only along a given tunnel is faster than computing all possible exit path, due to the decreased volume of the searched configuration space.
Another advantage of our method, in comparison with the existing approaches, is that we enable the users to trace the ligand path leading from the
outer solvent towards the protein active site.

@@

\section{Related Work}
{\color{red}
Migration of the ligands into a protein can be directly evaluated using Molecular dynamics (MD) simulations, that consider both 
the protein and the ligand flexible.
However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
for particular ligand/protein pair, which is not practical for testing a large number of different ligands.
Therefore, computational approaches that consider only geometry of the protein/ligand are still preferable due to their high speed.

The widely used approaches to evaluate possiblity of ligand migration into the protein rely on the tunnel detection.
Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing the protein-ligand interaction possibilities.
The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations.
The protein conformations can be prepared again using MD simulations.
Computationally less costly approaches for generating protein conformations are based on constrained geometric simulations like
ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained} and tCONCOORD~\cite{seeliger2007geometry}.
Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.

The main disadvantage of both grid-based and VD-based tunnel detection methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
The ligand flexibility cannot be principally considered by Voronoi-based approaches.
}

@@

The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF, that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using the electrostatic energy.
The high-energy configurations are discarded, while the low-energy conformations (and their vicinity) are sampled densely.
The approach~\cite{singhLIG} can also predict the ligand binding sites.

@@

\section{Proposed Method}
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from outside the protein to the active site.
The tunnels are detected on selected protein conformations resulting e.g. from MD simulations or from tool providing conformation sampling
~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
%The tunnels are detected in a sequence of time frames of MD simulations.
%capturing protein dynamic behavior. 
%The frames are computed using MD simulations.

In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows the biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
space, similarly as in~\cite{denny2018general}.
Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
The candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g. the geometric center or a selected atom), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
We assume that the ligand can change from conformation $l_1$ at configuration $q_1$ to a conformation $l_2$ at configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is less than the resolution of the planner ($\varepsilon=0.05$~\AA).

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it is added to $\QI$ as a new starting configuration.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
The distance $\RI$ influences how far the initial configurations are from the start of the tunnel.
By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
Contrary, too low $\RI$ may disable to find any collision-free configurations, especially for large ligands.
We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
can be found around.
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
The tree can slightly detour from the tunnel and overcome difficult parts (e.g., the tunnel bottleneck) by finding an alternative trajectory nearby.
%It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
%should be allowed, which is controlled by the $\dt$ parameter.
%parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in narrow void space.
We expand the tree considering all $\L$ conformations.
%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in the case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
smaller scales are used only in difficult (narrow) regions.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can easily rotate on the spot.
%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
%packed conformations can be used.
%\vspace{-3mm}

The pool of available conformations $\L$ consists all conformations (up to some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The task is to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (DhaAwt, PDB ID 4E46) which consists of 4650 atoms.
MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
The tunnels were detected in 100 consecutive frames of MD simulation in part where the protein was stabilized.
The tunnels for probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
computed using the CAVER 3.0~\cite{caver3} tool.
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.

Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075)

Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each initial configuration, 10 trajectories were computed, which resulted
in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
%The \RA\ method was also tested without the atom-based metric.
%In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.

@@

The performance of the methods is evaluated using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
This distance is measured as 3D distance between the geometric center of the ligand and the position of the active site.
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

\subsection{Analysis of protein conformations}

>>>>

The reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates} (see Fig.~\ref{fig::motiv}).
The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between the protein and ligand.
This can help them to reduce the number of necessary in-vitro experiments.
The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
i.e., they approximate the ligand by a bounding sphere.
However, ligands are typically of a non-spherical shape, which is completely ignored by these methods.

@@

Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
This necessarily requires to consider also conformational changes of the ligand, i.e., changes of the mutual position of the ligand atoms.
Such ligand is denoted as a flexible ligand.
%Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
%\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
%Such a ligand, for which the mutual position of its atoms can change, is denoted as a flexible ligand.

@@

Finding trajectories only along a given tunnel is faster than computing all possible exit paths, due to the decreased volume of the searched configuration space.
Another advantage of our method, in comparison with the existing approaches, is that we enable the users to trace the ligand path leading from the
outer solvent towards the protein active site.

@@

\section{Related Work}
Transportation of ligands into the protein active site can be directly evaluated using molecular dynamics (MD) simulations, where both the protein and ligand are flexible.
However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
for a particular ligand/protein pair, which is not practical for testing a large number of different ligands interacting with the protein.
Therefore, computational approaches that consider only the static geometry of the ligand are still preferable due to their high speed.

The widely used approaches to evaluate the ligand traversability rely on the tunnel detection.
Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3,sehnal2013mole}.
Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing of the feasibility of the protein-ligand interactions.
The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations, generated by MD simulations.
Computationally less demanding approaches for generating protein conformations are based on constrained geometric simulations, such as
ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained}, and tCONCOORD~\cite{seeliger2007geometry}.
Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.

The main disadvantage of Voronoi diagram based tunnel detection methods is that the shape of the ligand is not taken into account during the tunnel detection.
Therefore, it is not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
The ligand flexibility cannot be principally considered by Voronoi-based approaches and other techniques have to be adopted to solve this problem.

@@

The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOFs, that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using the electrostatic energy.
The high-energy configurations are discarded, while the low-energy configurations (and their vicinity) are sampled densely.
The approach~\cite{singhLIG} can also predict the ligand binding sites.

@@

\section{Proposed Method}
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from the protein surface to the inner buried active site.
The tunnels are detected on selected protein conformations, resulting, e.g., from MD simulations or from a tool providing the conformation sampling
~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
%The tunnels are detected in a sequence of time frames of MD simulations.
%capturing protein dynamic behavior. 
%The frames are computed using MD simulations.

In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately, which allows the biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance. 
This helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
space, similarly as in~\cite{denny2018general}.
Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels can adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
Protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ their 3D position, and $r_i > 0$ the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
The candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution, we use the discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., the geometric center or a selected atom), rotation around $x$, $y$, and $z$ axes,
index of the conformation $l\in \L$, and scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with protein atoms.
We assume that the ligand can change from conformation $l_1$ in configuration $q_1$ to conformation $l_2$ in configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is smaller than the resolution of the planner ($\varepsilon=0.05$~\AA).

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter the tunnel in various rotations and spatial conformations.
Therefore, the traversability of the tunnel should be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it is added to $\QI$ as new starting configuration.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
The distance $\RI$ influences how far the initial configurations are from the start of the tunnel.
By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
Contrary, too low $\RI$ may prevent the algorithm to find any collision-free configurations, especially for large ligands.
We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
can be found around.
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
The tree can slightly detour from the tunnel and overcome difficult parts (e.g., the tunnel bottleneck) by finding an alternative trajectory nearby.
%It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
%should be allowed, which is controlled by the $\dt$ parameter.
%parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
In case of tunnels not having other tunnels in the close vicinity (which is very common), we recommend to use the double value of the tunnel average width.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
We expand the tree considering all $\L$ conformations.
%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
smaller scales are used only in difficult (narrow) regions.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in the narrow passage, while the packed conformations can easily rotate on the spot.
%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
%packed conformations can be used.
%\vspace{-3mm}

The pool of available conformations $\L$ consists of all conformations (up to some resolution) with the energy lower than a user-specified threshold, which can still contain too many conformations (thousands or even more).
We further select a subset by computing how the atoms of the ligand are spread around its geometric center.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the elongated conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The task is to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (DhaAwt, PDB ID 4E46), which consists of 4650 atoms.
The MD simulation was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
The tunnels were detected in 100 consecutive frames of MD simulation in the part where the protein was already stabilized.
The tunnels for probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
computed using the CAVER 3.0~\cite{caver3} tool.
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels were $1.48$~\AA\ and $1.2$~\AA\ for the main and side tunnel, respectively.

Models  of the ligands in Mol2 formats were converted to the param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each initial configuration, 10 trajectories were computed, which resulted
in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was launched with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners, as it resulted in the highest success rates.
%The \RA\ method was also tested without the atom-based metric.
%In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.

@@

The performance of the methods is evaluated using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
This distance is measured as the 3D distance between the geometric center of the ligand and the position of the active site.
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

\subsection{Analysis of Protein Conformations}
